Pipit: Reactive Systems in \fstar ~(Extended Abstract)

### 2023-05-12 2023pipit/Main.tex
address page limit

%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
\documentclass[sigplan,screen]{acmart}

@@

\begin{abstract}
  Reactive languages such as Lustre and Scade are used to implement safety-critical control systems; proving such programs correct and having the proved properties apply to the compiled code is therefore equally critical.
  We introduce Pipit, a simple reactive language embedded in \fstar{}, designed for verifying control systems and executing them in real-time.
  Pipit includes a verified translation to transition systems; by reusing \fstar{}'s existing proof automation, certain safety properties can be automatically proved using k-induction on the transition system.
  Pipit can also generate imperative code in \lowstar{}, a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
  This translation to imperative code has been proved to preserve types; the proof that the generated imperative code preserves semantics is future work.
\end{abstract}

@@

To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
In the following, we introduce Pipit, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
Pipit aims to provide a high-level language based on a clock-free subset of Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers, and using \lowstar{}'s C-code generation for real-time execution.
% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
Pipit translates its expression language to a transition system for k-inductive proofs, which is verified\footnote{The translation is verified for a previous untyped expression representation; we are still finishing porting the proofs to the new typed representation.}; verifying the translation to imperative code is future work.

In this extended abstract, we will briefly describe the following preliminary results:

\begin{itemize}
  \item embedded language reduces the need for fancy features (\autoref{s:tut});
  \item normalisation and proof automation allow us to prove certain properties with minimal effort (\autoref{ss:tut:bibo});
  \item normalisation and extraction allows us to generate unverified executable code (\autoref{s:extraction});
\end{itemize}

We will also touch upon the following work-in-progress, which we intend to publish upon completion:

\begin{itemize}
  \item we describe the syntax (\autoref{s:core});
  \item a unified semantics for theorem proving and computation;
  \item verified imperative code generation;
  \item the ability to mix manual proofs with automated proofs
  \item further evaluation with the verification of a realistic controller
\end{itemize}

@@

% \section{Pipit in flight}
\pagebreak
\section{Programming and verifying in Pipit}
\label{s:tut}

A common requirement in controllers is to filter an input signal, perhaps using a \emph{finite impulse response} (FIR) filter, which is equivalent to a weighted moving average.
An FIR filter takes an input signal and a vector of coefficients, and computes the dot product of the coefficients and the most recent values of the signal.
We can implement an FIR filter in Pipit as follows:

\begin{tabbing}
  @MM@\= @MMMMMM@ \= \kill
  @let@ fir ($\textit{coefficients}$: list $\RR$) ($s$: stream $\RR$): stream $\RR$ = \\
    \> @match@ $\textit{coefficients}$ @with@ \\
    \> @|@ $[]$ \> $\to 0$ \\
    \> @|@ $c :: \cs $ \> $\to (s \cdot c) + (\xfby{0}{(\mbox{fir~} \cs~s)})$
\end{tabbing}

Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
The coefficient vector is represented by a list of reals here, while the signal $s$ is a \emph{stream} of reals, and the result is the filtered stream\footnote{The actual implementation uses integers with fixed-point arithmetic, but we use reals here to reduce clutter.}.
The implementation starts by looking at the list of coefficients, and returns zero if the list is empty.
If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($s \cdot c$) and delay the result ($\xfby{0}{\ldots}$) of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ s$), summing the two parts.
When there is no previous value, as is the case at the start of execution, the delay is given the initial value of zero.

Although this definition is simple, verifying it \emph{and} executing it in Lustre is somewhat impractical, as tool support varies widely.
To write this program in Lustre, we would likely need to reformulate the program to use arrays.
Vélus does not support arrays; Kind2 uses a custom syntax for arrays that no compiler supports; and Lustre v6 does support arrays, but its model-checker Lesar~\cite{raymond2008synchronous} does not.

Instead, we take the embedded language approach of defining a simple core language inside a theorem prover (\fstar{}), inspired by Bedrock~\cite{chlipala2013bedrock}.
By implementing Pipit as an embedded language, we can express the above FIR filter, but the lists are evaluated away by the meta-language to a simple core program which can be easily executed in statically-bounded runtime and space.
For example, we can compute the weighted moving average with two coefficients:

@@

Which then normalises to the following simple definition:

@@

To state properties about our filter, it is useful to be able to state temporal properties.
We do not need a separate specification language; instead, we can implement computable variants of temporal operators from \emph{past-time} linear temporal logic \cite{halbwachs1993executable,lichtenstein1985glory}.
We name the past-globally operator \emph{sofar}, as in \emph{the predicate has been true so far}:

\begin{tabbing}
  @MM@\= @MMMMMM@ \= \kill
  @let@ sofar ($p$: stream $\BB$): stream $\BB$ = \\
    \> $\xrec{p'}{p \wedge (\xfby{\top}{p'})}$
\end{tabbing}

This definition takes a stream of predicates $p$ and introduces a recursive stream $p'$.
At each step, the recursive stream $p'$ checks that the current predicate is true ($p$), and also checks that \emph{sofar} was previously true ($\xfby{\top}{p'}$).
If there is no previous value, it defaults to true.

@@

\begin{tabbing}
  @MM@\= @LET OUTPUT@\= @EXAMPLE1 INPUT@ \= \kill
  @let@ $\mbox{bibo}_2$ (n: $\RRgez$) ($\ii$: stream $\RR$): stream $\RR$ = \\
  \> @check@ $(\mbox{sofar}(|\ii| \le n)) \implies |\mbox{fir}_2~\ii| \le n)$
\end{tabbing}

This property states that, for this particular set of coefficients, if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
Note that the upper bound $n$ is a nonnegative real rather than a stream of reals, which means that $n$ stays constant across the whole stream.
To prove that this property holds, we translate to a transition system.
In this case, simple induction over the transition relation is sufficient to prove the property.
There is some boilerplate required to perform the induction, but both base and step cases are automatically proved by \fstar{}:

@@

If we wish to prove a similar BIBO property for more coefficients, standard induction over the transition system is no longer sufficient; we need a stronger invariant.
One simple automated way to strengthen invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which adds more context by assuming that the property holds for $k$ previous steps of the transition relation.
For example, we can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$:

@@

Although the properties we have shown here boil down to simple properties about linear arithmetic, we believe that this example demonstrates a promising way to use \fstar{}'s existing proof automation for verifying reactive systems.

\pagebreak
\section{Core language}
\label{s:core}

@@

Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
% To ensure that streams are productive,
As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
We use this form to specify an operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \citet{caspi1995functional} which is not syntax directed.
The syntax-directed semantics simplifies the proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax at the expense of potentially duplicating expressions.

@@

To demonstrate the feasibility of Pipit, we have implemented and verified a simple controller.
This system controls a water-flow solenoid to fill the reservoir of a coffee machine with multiple safeguards to reduce the risk of flooding.
The controller has two boolean inputs: the \emph{stop} switch and the \emph{low level} indicator; it returns a boolean indicating whether to engage the solenoid.
The stop switch indicates whether the reservoir's lid is open or closed; the system should never operate while the lid is open as water could spill out.
The second safeguard is that the solenoid should not be engaged for more than a minute or so as this may indicate a leak; if so, the controller enters a terminal \emph{stuck} state.
Finally, to avoid banging the solenoid too often, the controller waits for ten seconds of low water level before trying to engage.

\newcommand\estop{\textit{stop}}
\newcommand\low{\textit{low}}
\newcommand\soltry{\textit{try}}
\newcommand\stuck{\textit{stuck}}
\newcommand\solen{\textit{engage}}
\begin{tabbing}
  @MM@\= @let@ engage \= @MMMM@ \= \kill
  @let@ $\mbox{reservoir}$ (\estop{} \low{}: stream $\BB$): stream $\BB$ = \\
  \> @let@ \soltry{} \> = true\_for \textit{TEN\_SECONDS} (not \estop{} $\wedge$ \low{}) @in@ \\
  \> @let@ \stuck{} \> = once (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
  \> @let@ \solen{} \> = \soltry{} $\wedge$ not \stuck{} @in@ \\
  \> @check@ (\solen{} $\implies$ not \estop{}); \\
  \> @check@ (\solen{} $\implies$ \low{}); \\
  \> \solen{}
\end{tabbing}

As with the examples in \autoref{ss:tut:bibo}, the above properties can be inductively verified by \fstar{}'s proof automation.
We can also generate executable C code, which runs in real-time on a microcontroller in the first author's kitchen.

@@

Earlier work verifying reactive systems used model-checking via explicit state-space exploration~\cite{raymond2008synchronous}; more recent work uses SMT solvers to find inductive proofs~\cite{hagen2008scaling,champion2016kind2} or check refinement types~\cite{chen2022synchronous}.
These model-checkers have definite advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples that disprove properties; they can also implement advanced invariant generation techniques such as ICE~\cite{garg2014ice} and IC3/PDR~\cite{bradley2011sat,een2011efficient}.
However, these special-purpose model-checkers do not provide much assurance that the semantics they use for proofs matches the compiled code.
We believe that once Pipit's imperative code generation is verified, Pipit will have a stronger assurance case.

@@

We believe that these preliminary results show promise.
In the intermediate future, we intend to verify the imperative code generation.

TODO
Future work: verify imperative code generation; verify CSE; case studies: antilock braking; clocks \cite{caspi1995functional}; contracts \cite{champion2016kind2}.

>>>>

%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
\documentclass[sigplan,screen, review]{acmart}

@@

\begin{abstract}
  Reactive languages such as Lustre and Scade are used to implement safety-critical control systems; proving such programs correct and having the proved properties apply to the compiled code is therefore equally critical.
  We introduce Pipit, a small reactive language embedded in \fstar{}, designed for verifying control systems and executing them in real-time.
  Pipit includes a verified translation to transition systems; by reusing \fstar{}'s existing proof automation, certain safety properties can be automatically proved by k-induction on the transition system.
  Pipit can also generate imperative code in \lowstar{}, a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
  This translation to imperative code preserves types by construction; the proof that the imperative code preserves semantics is ongoing.
\end{abstract}

@@

To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
In this abstract we introduce Pipit\footnote{Implementation available at \url{https://github.com/songlarknet/pipit}}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
Pipit aims to provide a high-level language based on a clock-free subset of Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers, and using \lowstar{}'s C-code generation for real-time execution.
% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
Pipit translates its expression language to a transition system for k-inductive proofs, which is verified\footnote{The translation is verified for a previous untyped expression representation; we are still finishing porting the proofs to the new typed representation.}; verifying the translation to imperative code is ongoing.

In this extended abstract we briefly describe the following preliminary results, which we intend to describe fully in a future publication:

\begin{itemize}
  \item we motivate Pipit which, as an embedded language, provides syntactic convenience with a small verifiable core language (\autoref{s:tut});
  \item we demonstrate the use of \fstar{}'s existing normalisation and proof automation to prove certain properties with minimal effort (\autoref{ss:tut:bibo});
  \item we describe a key difference between Pipit's core language and Lustre (\autoref{s:core}); and
  % \item we demonstrate the use of normalisation, tactics and \lowstar{} extraction to generate unverified executable code (\autoref{s:extraction}); and
  \item we evaluate Pipit by executing a verified controller on an embedded system (\autoref{s:coffee-machine}).
\end{itemize}

@@

% \section{Pipit in flight}
% \pagebreak
\section{Programming and verifying in Pipit}
\label{s:tut}

A common requirement in controllers is to filter an input signal, perhaps using a \emph{finite impulse response} (FIR) filter, which is equivalent to a weighted moving average.
An FIR filter takes a vector of coefficients and an input signal; at every point in time, it computes the dot product of the coefficients and the most recent values of the signal.
We can implement an FIR filter in Pipit as follows:

\newcommand\signal{\textit{signal}}
\begin{tabbing}
  @MM@\= @MMMMMM@ \= \kill
  @let@ fir ($\textit{coefficients}$: list $\RR$) ($\signal$: stream $\RR$): stream $\RR$ = \\
    \> @match@ $\textit{coefficients}$ @with@ \\
    \> @|@ $[]$ \> $\to 0$ \\
    \> @|@ $c :: \cs $ \> $\to (\signal \cdot c) + (\xfby{0}{(\mbox{fir~} \cs~\signal)})$
\end{tabbing}

% Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
The coefficient vector is represented by a list of reals here, while the signal is a \emph{stream} of reals, and the result is the filtered stream.
% \footnote{The actual implementation uses integers with fixed-point arithmetic, but we use reals here to reduce clutter.}
The implementation starts by looking at the list of coefficients, and returns zero if the list is empty.
If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($\signal \cdot c$); we also take the result of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ \signal$) and delay it ($\xfby{0}{\ldots}$) before finally summing the two parts.
At the start of execution the delay is initialised to zero.

Although this is a simple definition, verifying it \emph{and} executing it in Lustre is somewhat impractical, as tool support varies widely.
To write this program in Lustre, we would likely need to reformulate the program to use arrays.
Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays that no compiler supports~\cite{champion2016kind2}; and Lustre v6 does support arrays, but its model-checker Lesar does not~\cite{raymond2008synchronous}.

Instead, inspired by Bedrock~\cite{chlipala2013bedrock}, we define Pipit with a small core language embedded inside a theorem prover (\fstar{} in our case).
By implementing Pipit as an embedded language, we can express the above FIR filter, but the lists are evaluated away by the meta-language to a simple core program which can be easily executed in statically-bounded runtime and space.
For example, we can easily compute the weighted moving average with concrete coefficients:

@@

Normalising the above definition of $\mbox{fir}_2$ removes all lists.
The result fits in the core language defined in \autoref{s:core}, for which we can generate statically-allocated imperative code:

@@

The properties that we want to state about reactive programs usually involve some temporal aspect.
Rather than defining a separate specification language, we implement computable variants of temporal operators from \emph{past-time} linear temporal logic \cite{halbwachs1993executable,lichtenstein1985glory}.
We name the past-globally operator \emph{sofar}, as in \emph{the predicate has been true so far}:

\begin{tabbing}
  @MM@\= @MMMMMM@ \= \kill
  @let@ sofar ($p$: stream $\BB$): stream $\BB$ = \\
    \> $\xrec{p'}{p \wedge (\xfby{@true@}{p'})}$
\end{tabbing}

This definition takes a stream of predicates $p$ and introduces a recursive stream $p'$.
At each step, the recursive stream $p'$ checks that the current predicate is true ($p$), and also checks that \emph{sofar} was previously true ($\xfby{@true@}{p'}$).
If there is no previous value, it defaults to true.

@@

\begin{tabbing}
  @MM@\= @LET OUTPUT@\= @EXAMPLE1 INPUT@ \= \kill
  @let@ $\mbox{bibo}_2$ (n: $\RRgez$) ($\ii$: stream $\RR$): stream $\RR$ = \\
  \> @check@ $(\mbox{sofar}(|\ii| \le n) \implies |\mbox{fir}_2~\ii| \le n)$
\end{tabbing}

This property states that for the coefficients used by $\fir_2$, if the input has always been in the range $[-n, n]$, then the output is also within the range $[-n, n]$.
Note that the upper bound $n$ is a nonnegative real rather than a stream of reals, which means that $n$ stays constant across the whole stream.
To prove that this property holds, we translate to a transition system.
In this case, simple induction over the transition relation is sufficient to prove the property.
There is some boilerplate required to perform the induction, but both base and step cases are automatically proved by \fstar{}:

@@

If we wish to prove a similar BIBO property for a filter with more coefficients, standard induction over the transition system is no longer sufficient; we need a stronger invariant.
One simple automated way to strengthen invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which adds more context by assuming that the property holds for $k$ previous steps of the transition relation.
We can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$ as follows:

@@

Although the properties here boil down to simple properties about linear arithmetic, we believe that this example demonstrates a promising way to use \fstar{}'s existing proof automation for verifying reactive systems.

% \pagebreak
\section{Core language}
\label{s:core}

@@

Recursive streams, which can refer to previous values of the stream itself, are defined using the fixpoint operator ($\xrec{x}{e[x]}$); the syntax $e[x]$ means that the variable $x$ can occur in $e$.
% To ensure that streams are productive,
As in Lustre, recursive streams can only refer to their previous values and must be \emph{guarded} by a delay: the stream $(\xrec{x}{\xfby{0}{(x + 1)}})$ is well-defined, but stream $(\xrec{x}{x + 1})$ is invalid and has no computational interpretation.
This form of recursion differs slightly from standard Lustre, which uses a set of mutually-recursive bindings.
We use this form to specify a substitution-based operational semantics that is syntax-directed, as opposed to the mutually-recursive form in \citet{caspi1995functional} which is not syntax directed.
The syntax-directed semantics simplifies the proof of determinism; we believe it has simplified other necessary proofs too and will perform further evaluation.
Although we cannot express mutually-recursive bindings in the core syntax here, we can express them as a notation on the surface syntax at the expense of potentially duplicating expressions.

@@

To demonstrate the feasibility of Pipit, we have implemented and verified a simple controller.
This system controls a water-flow solenoid to fill the reservoir of a coffee machine and includes multiple safeguards to reduce the risk of flooding\footnote{For a video of the controller in action, see \url{https://youtu.be/6IybbQFPOl8}}.
The controller has two boolean inputs: the \emph{stop} switch and the \emph{low level} indicator; it returns a boolean indicating whether to engage the solenoid.
The stop switch indicates whether the reservoir's lid is open or closed; the system should never operate while the lid is open as water could spill out.
The controller should not allow water to flow for more than a minute as this may indicate a leak; if so, the controller enters a terminal error state.
Finally, to avoid switching the solenoid too often, the controller waits for ten seconds of low water level before trying to engage.

\newcommand\estop{\textit{stop}}
\newcommand\low{\textit{low}}
\newcommand\soltry{\textit{try}}
\newcommand\error{\textit{error}}
\newcommand\solen{\textit{engage}}
\begin{tabbing}
  @MM@\= @let@ engage \= @MMMM@ \= \kill
  @let@ $\mbox{reservoir}$ (\estop{} \low{}: stream $\BB$): stream $\BB$ = \\
  \> @let@ \soltry{} \> = true\_for \textit{TEN\_SECONDS} (not \estop{} $\wedge$ \low{}) @in@ \\
  \> @let@ \error{} \> = once (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
  \> @let@ \solen{} \> = \soltry{} $\wedge$ not \error{} @in@ \\
  \> @check@ (\solen{} $\implies$ not \estop{}); \\
  \> @check@ (\solen{} $\implies$ \low{}); \\
  \> \solen{}
\end{tabbing}

As with the examples in \autoref{ss:tut:bibo}, the above properties can be inductively verified by \fstar{}'s proof automation.
Pipit generates real-time C code for this example.

@@

Earlier work model-checking reactive systems used explicit state-space exploration~\cite{raymond2008synchronous}; more-recent work uses SMT solvers to find inductive proofs~\cite{hagen2008scaling,champion2016kind2} or to check refinement types~\cite{chen2022synchronous}.
These model-checkers have definite advantages over the general-purpose-prover approach offered here: they can often generate concrete counterexamples; they can also implement counterexample-based invariant-generation techniques such as ICE~\cite{garg2014ice} and PDR~\cite{bradley2011sat,een2011efficient}.
However, these model-checkers do not provide high assurance that the semantics they use for proofs matches the compiled code.
We believe that once Pipit's imperative code generation is verified, Pipit will have a stronger assurance case.

The embedded language Copilot generates real-time C code for runtime monitoring and supports model-checking properties \cite{laurent2015assuring}, but suffers from the same semantic gap.

@@

Our preliminary results show that \fstar{}'s proof automation and code extraction are suitable for verifying reactive systems and executing them in real-time; these results still require further work.
Next, we intend to verify the imperative code generation.
To prove large programs, we also need some way to compose smaller proofs together, such as contracts.
Finally, we need to evaluate Pipit on larger control systems before extending the language with clocks.

% TODO
% Future work: verify imperative code generation; verify CSE; case studies: antilock braking; clocks \cite{caspi1995functional}; contracts \cite{champion2016kind2}.
